### 2014-08-11 Studies/2011_San_Francisco/Report/SF_Fire.tex
San Francisco: Edits to tables and styling throughout

\chapter*{\centering Abstract}
The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on XXXX, within a 2~\sfrac{1}{2}~story, single family structure in San Francisco, CA, that resulted in the death of two firefighters. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at a height of 1.86~m (6~ft). Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. It appears that the captain was likely overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure, but later succumbed to his injuries.

\chapter{Introduction}
Part of the function of the Fire Research Division at NIST is to develop and apply technology, measurements, and standards to improve the understanding of the behavior, prevention, and control of fires to enhance fire fighting operations and equipment, fire suppression, fire investigations, and disaster response. NIST has previously used the Fire Dynamics Simulator (FDS) fire model to provide insight into the fire development and thermal conditions of fires that have resulted in line of duty deaths (LODDs)~\cite{Madrzykowski:1,Iowa,Texas,Bryner:Charleston,barowy:texas}. The objective of these studies has been to improve the safety and effectiveness of firefighters.

On November 2, 2012, a fire in a 2~\sfrac{1}{2}~story wood frame residential structure claimed the life of a captain of the Chicago Fire Department. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide}  software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{fire_sum} provides a summary of the fire incident, Section~\ref{model} describes the relevant model input parameters and assumptions that were used to develop the simulations, Section~\ref{results} presents the simulation results, and Section~\ref{discussion} discusses the simulation results as they relate to firefighter safety and effectiveness. Appendix~\ref{app_draw} contains dimensioned drawings of the first and second floor of the structure.

\chapter{Summary of the Fire Incident}
\label{fire_sum}
The account of events for this incident was documented by the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines a rapid intervention team (RIT) response to be one truck company, one battalion chief, one advanced life support (ALS) ambulance, and one emergency medical services (EMS) field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the [rear] door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a [hose] load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

@@

\begin{table}
\centering
\captionof{table}[Abridged NIOSH approximate fire event timeline]{Abridged NIOSH approximate fire event timeline~\cite{NIOSH:Bowyer}}\label{tab:fire_info}
\begin{tabular}{cl}
\toprule[1.5pt]
Incident Time              &  Fire Behavior / Fireground Operation                                                                                                    \\
{[hh:mm:ss]}               &                                                                                                                                          \\
\midrule
\multirow{2}{*}{17:15:00}  &  \multirow{2}{*}{\parbox{10cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}}        \\ 
                           &                                                                                                                                          \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00}  &   \multirow{2}{*}{\parbox{10cm} {Dispatch for a ``Still'' alarm - two engine companies, two truck companies, and a battalion chief.}}    \\
                           &                                                                                                                                          \\[.25cm]
\multirow{2}{*}{17:20:00}  &  \multirow{2}{*}{\parbox{10cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear.}}      \\
                           &                                                                                                                                          \\[.25cm] 
17:21:00                   &  E123 captain and firefighter on second floor with 1\sfrac{3}{4}-inch hoseline.                                                          \\[.25cm]
17:21:30                   &  First floor exterior porch door opened.                                                                                                 \\[.25cm]
\multirow{2}{*}{17:23:00}  &  \multirow{2}{*}{\parbox{10cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123 captain or firefighter.}}  \\
                           &                                                                                                                                          \\[.25cm]
17:24:00                   &   Water applied to attic from the rear exterior position.                                                                                \\[.25cm]
17:25:00                   &   SQ5 makes entry through front of structure.                                                                                            \\[.25cm]
17:27:00                   &   After hearing a Mayday call from SQ5, IC calls a Mayday.                                                                               \\[.25cm]
17:29:00                   &   E123 captain carried out of structure.                                                                                                 \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

\chapter{Model Description}
\label{model}
Fire Dynamics Simulator~\cite{FDS_Users_Guide} is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure, and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gasses. One of the most important advantages of FDS is that it is  mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data to ensure that it provides the expected results, given sufficient input data~\cite{FDS_Validation_Guide}. A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Math_Guide}.

@@

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2}~story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~\ref{fig:simp_geom}. This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber covered by 12.7~mm (0.5~in) gypsum board~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~\ref{fig:simp_geom}), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~\ref{fig:simp_geom}) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~\ref{fig:simp_geom}). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~\ref{fig:simp_geom}.

The fire, which is described in Section~\ref{fire}, started in the attic and spread to the rear porch. Discussion of the fire's impact on the interior conditions is provided in Section~\ref{results}. Figure~\ref{fig:alpha_ex} shows the exterior of the front and rear of the structure after the incident. Note that the majority of the damage occurred in the attic and rear porch on the second floor. Fully dimensioned drawings of the interior of the first and second floors are shown in Figs.~\ref{fig:first_floor} and \ref{fig:second_floor}, respectively.

% \begin{figure}[!ht]
% \begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
% \includegraphics[width=\textwidth]{../Figures/1st_Floor_Metric} \\
% \includegraphics[width=\textwidth]{../Figures/2nd_Floor_Metric}
% \end{tabular*}
% \caption[Plan view of first (top) and second (bottom) floor of structure.]{Plan view of first (top) and second (bottom) floor of structure. Stairs and key rooms are identified using measurements collected by NIST.}
% \label{fig:simp_geom}
% \end{figure}

% \begin{figure}[!ht]
% \centering
% \begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
%    \scalebox{1.0}{ \includegraphics[width=3in]{../Figures/exterior_alpha} } &
%    \scalebox{1.0}{ \includegraphics[width=2.67in]{../Figures/exterior_charlie} } \\
% \end{tabular*}
% \caption{Photographs showing the front (left) and rear (right) exterior of the structure after the incident.}
% \label{fig:alpha_ex}
% \end{figure}

@@

\section{Fire}
\label{fire}

@@

\section{Materials}
\label{matl}
While the fires were represented as fuel sources with a constant area (Section~\ref{fire}), from a heat transfer perspective, it is important to define the material properties (density, thermal conductivity, and specific heat) to account for heat transfer and energy storage in the ceiling, walls, and floor. In this study, the material properties of gypsum board~\cite{WAKILI2007} were specified on the finished walls and ceilings on the main two floors of the structure, and the material properties of wood~\cite{Incropera:1} were specified on the unfinished portions of the attic and rear porch. The following lines define these materials in the FDS input file:

@@

\section{Ventilation}
\label{Vents}
The simulations in this study account for changes in the ventilation due to a combination of fire department operations (opening doors) and fire acting on the structure (fire breaching walls, breaking windows, etc). Two different ventilation scenarios were considered in this study. The first scenario involves a baseline simulation in which the ventilation times and ventilation areas represent our best understanding of the incident. The second scenario, or alternative simulation, does not include the firefighter opening the rear door to make entry to the structure and the subsequent exterior left wall fire penetration. The times of the ventilation changes are provided in Table~\ref{tab:vents}. Structure leakage has been shown to be important when modeling enclosure fires~\cite{beal2009}. The open front door at the start of the simulation and the vents (opening size and open time) created during the simulation are significantly larger than the total effective area of leakage of a structure of this type. Therefore, any leakage in the structure would have a negligible impact on the fluid mechanics within the structure and is not included in this study.

@@

\begin{table}
\centering
\captionof{table}{Timeline of ventilation changes in the simulation}\label{tab:vents}
\begin{tabular}{cccll}
\toprule[1.5pt]
Baseline    &  Alternative  &  Ventilation  &  Type         &  Side of         \\
Simulation  &  Simulation   &  Area         &               &  Structure       \\
{[s]}       &  {[s]}        &  [m$^2$]      &               &                  \\
\midrule
0           &  0            &  0.11         &  Window       &  Rear            \\
40          &  40           &  0.08         &  Breach (x2)  &  Front and Rear  \\
45          &  45           &  0.12         &  Breach       &  Left            \\
55          &  55           &  0.12         &  Breach       &  Rear            \\
130         &  N/A$^*$      &  1.41         &  Door         &  Rear            \\
135         &  N/A$^*$      &  0.2          &  Breach       &  Left            \\
160         &  160          &  0.80         &  Door         &  Interior        \\
\bottomrule[1.25pt]
\end{tabular}\par
\footnotesize
$^*$These ventilation events were not included in the alternative simulation.
\normalsize
\end{table}

% \begin{figure}[!ht]
% \centering
% \includegraphics[width=.80\textwidth]{../Figures/door_frame_burn}
% \caption[Post-incident image of rear porch door frame.]{Post-incident image of the second floor porch door frame that failed taken from inside the structure looking toward the porch. Note that the hinge is still intact and there is unburned, ``clean'' wood along the bottom portion of the frame.}
% \label{fig:door_frame}
% \end{figure}

@@

\section{Numerical Mesh}
\label{mesh}

@@

As discussed in Section~\ref{geom}, the structure has dimensions of 17.15~m (56~ft) by 6.15~m (20~ft) and a total height of 8.56~m (28~ft). To ensure adequately resolved fluid flow in and out of the structure, the computational domain was extended beyond the volume of the structure. The total volume of the computational domain was 24~m (79~ft) by 10~m (33~ft) by 10~m (33~ft). A grid resolution of 10~cm (3.9~in) was used, which required 2.4 million computational cells. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas will snap to the nearest 10~cm. Figure~\ref{fig:geom_grid} shows the front of the structure rendered in Smokeview with the 10~cm mesh. The domain was divided into 16 equally sized meshes (each containing 150,000 grid cells) that could be processed in parallel, which reduced the amount of required calculation time to approximately 1.5~days. Figure~\ref{fig:mult_mesh} shows the entire structure within the computational domain that has been divided into multiple meshes. Table~\ref{tab:mod_param} shows a summary of the model input parameters for the simulations that were conducted as part of this study.

% \begin{figure}[!ht]
% \centering
% \begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
%    \scalebox{1.0}{\includegraphics[width=.49\textwidth]{../Figures/smv_exterior_grid}} &
%    \scalebox{1.0}{\includegraphics[width=.488\textwidth]{../Figures/smv_exterior_grid_rear}} \\
% \end{tabular*}
% \caption{Front (left) and rear (right) of the structure with 10~cm computational mesh.}
% \label{fig:geom_grid}
% \end{figure}

% \begin{figure}[!ht]
% \centering
% \includegraphics[width=.60\textwidth]{../Figures/smv_exterior_mesh}
% \caption[Front-right side of the structure within the computational domain.]{Front-right side of the structure within the computational domain (with multiple meshes). Each of the 16 meshes is 3~m (9.8~ft) by 10~m (33~ft) by 5~m (16.4~ft).}
% \label{fig:mult_mesh}
% \end{figure}

>>>>

\chapter*{\centering Abstract}

The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on XXXX, within a 2~\sfrac{1}{2}~story, single family structure in San Francisco, CA, that resulted in the death of two firefighters. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at a height of 1.86~m (6~ft). Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. It appears that the captain was likely overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure, but later succumbed to his injuries.

\chapter{Introduction}
\label{sec:introduction}

Part of the function of the Fire Research Division at NIST is to develop and apply technology, measurements, and standards to improve the understanding of the behavior, prevention, and control of fires to enhance fire fighting operations and equipment, fire suppression, fire investigations, and disaster response. NIST has previously used the Fire Dynamics Simulator (FDS) fire model to provide insight into the fire development and thermal conditions of fires that have resulted in line of duty deaths (LODDs)~\cite{Madrzykowski:1,Iowa,Texas,Bryner:Charleston,barowy:texas}. The objective of these studies has been to improve the safety and effectiveness of firefighters.

On November 2, 2012, a fire in a 2~\sfrac{1}{2}~story wood frame residential structure claimed the life of a captain of the Chicago Fire Department. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide}  software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{sec:incident_summary} provides a summary of the fire incident, Section~\ref{sec:model_description} describes the relevant model input parameters and assumptions that were used to develop the simulations, Section~\ref{sec:model_results} presents the simulation results, and Section~\ref{sec:discussion} discusses the simulation results as they relate to firefighter safety and effectiveness. Appendix~\ref{sec:drawings} contains dimensioned drawings of the first and second floor of the structure.

\chapter{Summary of the Fire Incident}
\label{sec:incident_summary}

The account of events for this incident was documented by the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines a rapid intervention team (RIT) response to be one truck company, one battalion chief, one advanced life support (ALS) ambulance, and one emergency medical services (EMS) field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the [rear] door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a [hose] load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

@@

\begin{table}[!ht]
\caption[Abridged NIOSH approximate fire event timeline]{Abridged NIOSH approximate fire event timeline~\cite{NIOSH:Bowyer}}
\begin{tabular}{cl}
\toprule
Incident Time  &  Fire Behavior / Fireground Operation                       \\
{(hh:mm:ss)}   &                                                             \\
\midrule
17:15:00       &  Reported working fire with heavy smoke coming from the     \\ 
               &  front and rear of the structure's attic.                   \\
               &                                                             \\
17:16:00       &  Dispatch for a ``Still'' alarm - two engine companies,     \\
               &  two truck companies, and a battalion chief.                \\
               &                                                             \\
17:20:00       &  Attic windows break - heavy smoke out of front attic       \\
               &  window and heavy smoke and fire out of rear.               \\
               &                                                             \\
17:21:00       &  E123 captain and firefighter on second floor               \\
               &  with 1\sfrac{3}{4}-inch hoseline.                          \\
               &                                                             \\
17:21:30       &  First floor exterior porch door opened.                    \\
               &                                                             \\
17:23:00       &  Heavy fire report in rear stairwell and enclosed porch.    \\
               &  No response from E123 captain or firefighter.              \\
               &                                                             \\
17:24:00       &  Water applied to attic from the rear exterior position.    \\
               &                                                             \\
17:25:00       &  SQ5 makes entry through front of structure.                \\
               &                                                             \\
17:27:00       &  After hearing a Mayday call from SQ5, IC calls a Mayday.   \\
               &                                                             \\
17:29:00       &  E123 captain carried out of structure.                     \\
\bottomrule
\end{tabular}
\label{tab:fire_info}
\end{table}

\chapter{Model Description}
\label{sec:model_description}

Fire Dynamics Simulator~\cite{FDS_Users_Guide} is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure, and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gasses. One of the most important advantages of FDS is that it is  mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data to ensure that it provides the expected results, given sufficient input data~\cite{FDS_Validation_Guide}. A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Math_Guide}.

@@

\section{Geometry}
\label{sec:geometry}

The structure involved in this fire incident was a 2~\sfrac{1}{2}~story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~\ref{fig:simp_geom}. This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber covered by 12.7~mm (0.5~in) gypsum board~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~\ref{fig:simp_geom}), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~\ref{fig:simp_geom}) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~\ref{fig:simp_geom}). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~\ref{fig:simp_geom}.

The fire, which is described in Section~\ref{sec:fire}, started in the attic and spread to the rear porch. Discussion of the fire's impact on the interior conditions is provided in Section~\ref{sec:model_results}. Figure~\ref{fig:alpha_ex} shows the exterior of the front and rear of the structure after the incident. Note that the majority of the damage occurred in the attic and rear porch on the second floor. Fully dimensioned drawings of the interior of the first and second floors are shown in Figs.~\ref{fig:first_floor} and \ref{fig:second_floor}, respectively.

% \begin{figure}[!ht]
% \includegraphics[width=in]{../Figures/1st_Floor_Metric} \\
% \includegraphics[width=in]{../Figures/2nd_Floor_Metric}
% \caption[Plan view of first (top) and second (bottom) floor of structure.]{Plan view of first (top) and second (bottom) floor of structure. Stairs and key rooms are identified using measurements collected by NIST.}
% \label{fig:simp_geom}
% \end{figure}

% \begin{figure}[!ht]
% \includegraphics[width=3in]{../Figures/exterior_alpha} \\
% \includegraphics[width=2.67in]{../Figures/exterior_charlie}
% \caption{Photographs showing the front (left) and rear (right) exterior of the structure after the incident.}
% \label{fig:alpha_ex}
% \end{figure}

@@

\section{Fire}
\label{sec:fire}

@@

\section{Materials}
\label{sec:materials}

While the fires were represented as fuel sources with a constant area (Section~\ref{sec:fire}), from a heat transfer perspective, it is important to define the material properties (density, thermal conductivity, and specific heat) to account for heat transfer and energy storage in the ceiling, walls, and floor. In this study, the material properties of gypsum board~\cite{WAKILI2007} were specified on the finished walls and ceilings on the main two floors of the structure, and the material properties of wood~\cite{Incropera:1} were specified on the unfinished portions of the attic and rear porch. The following lines define these materials in the FDS input file:

@@

\section{Ventilation}
\label{sec:ventilation}

The simulations in this study account for changes in the ventilation due to a combination of fire department operations (opening doors) and fire acting on the structure (fire breaching walls, breaking windows, etc). Two different ventilation scenarios were considered in this study. The first scenario involves a baseline simulation in which the ventilation times and ventilation areas represent our best understanding of the incident. The second scenario, or alternative simulation, does not include the firefighter opening the rear door to make entry to the structure and the subsequent exterior left wall fire penetration. The times of the ventilation changes are provided in Table~\ref{tab:vents}. Structure leakage has been shown to be important when modeling enclosure fires~\cite{beal2009}. The open front door at the start of the simulation and the vents (opening size and open time) created during the simulation are significantly larger than the total effective area of leakage of a structure of this type. Therefore, any leakage in the structure would have a negligible impact on the fluid mechanics within the structure and is not included in this study.

@@

\begin{table}[!ht]
\caption[Timeline of ventilation changes in the simulation]{Timeline of ventilation changes in the simulation}
\begin{tabular}{cccll}
\toprule
Baseline    &  Alternative  &  Ventilation  &  Type         &  Side of         \\
Simulation  &  Simulation   &  Area         &               &  Structure       \\
{[s]}       &  {[s]}        &  [m$^2$]      &               &                  \\
\midrule
0           &  0            &  0.11         &  Window       &  Rear            \\
40          &  40           &  0.08         &  Breach (x2)  &  Front and Rear  \\
45          &  45           &  0.12         &  Breach       &  Left            \\
55          &  55           &  0.12         &  Breach       &  Rear            \\
130         &  N/A$^*$      &  1.41         &  Door         &  Rear            \\
135         &  N/A$^*$      &  0.2          &  Breach       &  Left            \\
160         &  160          &  0.80         &  Door         &  Interior        \\
\bottomrule
\end{tabular}
\footnotesize
\\$^*$These ventilation events were not included in the alternative simulation.
\normalsize
\label{tab:vents}
\end{table}

% \begin{figure}[!ht]
% \includegraphics[width=.80in]{../Figures/door_frame_burn}
% \caption[Post-incident image of rear porch door frame.]{Post-incident image of the second floor porch door frame that failed taken from inside the structure looking toward the porch. Note that the hinge is still intact and there is unburned, ``clean'' wood along the bottom portion of the frame.}
% \label{fig:door_frame}
% \end{figure}

@@

\section{Numerical Mesh}
\label{sec:numerical_mesh}

@@

As discussed in Section~\ref{sec:geometry}, the structure has dimensions of 17.15~m (56~ft) by 6.15~m (20~ft) and a total height of 8.56~m (28~ft). To ensure adequately resolved fluid flow in and out of the structure, the computational domain was extended beyond the volume of the structure. The total volume of the computational domain was 24~m (79~ft) by 10~m (33~ft) by 10~m (33~ft). A grid resolution of 10~cm (3.9~in) was used, which required 2.4 million computational cells. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas will snap to the nearest 10~cm. Figure~\ref{fig:geom_grid} shows the front of the structure rendered in Smokeview with the 10~cm mesh. The domain was divided into 16 equally sized meshes (each containing 150,000 grid cells) that could be processed in parallel, which reduced the amount of required calculation time to approximately 1.5~days. Figure~\ref{fig:mult_mesh} shows the entire structure within the computational domain that has been divided into multiple meshes. Table~\ref{tab:mod_param} shows a summary of the model input parameters for the simulations that were conducted as part of this study.

% \begin{figure}[!ht]
% \includegraphics[width=.49in]{../Figures/smv_exterior_grid} \\
% \includegraphics[width=.488in]{../Figures/smv_exterior_grid_rear}
% \caption{Front (left) and rear (right) of the structure with 10~cm computational mesh.}
% \label{fig:geom_grid}
% \end{figure}

% \begin{figure}[!ht]
% \includegraphics[width=.60in]{../Figures/smv_exterior_mesh}
% \caption[Front-right side of the structure within the computational domain.]
% {Front-right side of the structure within the computational domain (with multiple meshes). Each of the 16 meshes is 3~m (9.8~ft) by 10~m (33~ft) by 5~m (16.4~ft).}
% \label{fig:mult_mesh}
% \end{figure}

\begin{table}[!ht]
\caption[Relevant fire model input parameters]{Relevant fire model input parameters}
\begin{tabular}{lll}
\toprule
Parameter                                                  & Description                                   & Discussion                                     \\
\midrule
Simulation Time                                            &  5~min                                        & --                                             \\ [.25cm]
Grid Cell Size                                             &  10~cm                                        & Section~\ref{sec:numerical_mesh}               \\ [.25cm]
Ambient Temperature$^*$                                    &  20~$^{\circ}$C (77~$^{\circ}$F)              & --                                             \\ [.1cm]
\multirow{4}{*}{Reaction: Wood~\cite{SFPE:Tewarson}}       &  Formula: $\C\H_{1.7}\O_{0.74}\N_{0.002}$     & \multirow{4}{*}{Section~\ref{sec:fire}}        \\
                                                           &  CO Yield: 0.004~$\mathrm{kg}/\mathrm{kg}$    &                                                \\
                                                           &  Soot Yield: 0.015~$\mathrm{kg}/\mathrm{kg}$  &                                                \\
                                                           &  Heat of Combustion: 16,400~kJ/kg             &                                                \\ [.25cm]
\multirow{2}{*}{Peak HRR}                                  &  Attic Fire: 9~MW                             & \multirow{2}{*}{Section~\ref{sec:fire}}        \\ 
                                                           &  Porch Fire: 3.3~MW                           &                                                \\ [.25cm]                     
\multirow{3}{*}{Material: Wood~\cite{Incropera:1} }        &  $k$: 0.12~\si{W/(m.K)})                      & \multirow{3}{*}{Section~\ref{sec:materials}}   \\
                                                           &  $\rho$: 510~kg/m$^3$                         &                                                \\
                                                           &  $c_{p}$: 1.38~\si{kJ/(kg.K)}                 &                                                \\ [.25cm]
\multirow{3}{*}{Material: Gypsum Board~\cite{WAKILI2007}}  &  $k$: 0.28~\si{W/(m.K)}                       &  \multirow{3}{*}{Section~\ref{sec:materials}}  \\ 
                                                           &  $\rho$: 810~kg/m$^3$                         &                                                \\ 
                                                           &  $c_{p}$: 1.0~\si{kJ/(kg.K)}                  &                                                \\
\bottomrule
\end{tabular}
\footnotesize
\\ $^{*}$ Initial interior temperatures were assumed to be 20~$^{\circ}$C; exterior temperatures were 7~$^{\circ}$C~\cite{NIOSH:Bowyer}.
\normalsize
\label{tab:mod_param}
\end{table}

\chapter{Model Results}
\label{sec:model_results}
